\vspace*{-8mm}\textbf{\textsc{Novelty Detection for Semantic Place Categorization}}

### 2011-06-23 thesis/summary-en.tex
Corrections in the summary-en

\section{Motivation}
For a long time humanity has fantasized that one day robots will walk among us.
They will move and be able to interact with us, understand our concepts and
be able to reason.
An important step on that direction is endowing robots with knowledge of human concepts and
semantics. However its unrealistic to believe on the possibility of extensively
describing human concepts and semantics to robots on a pre-deploy phase, and so, robots must be able
to adapt and learn when confronted with novel situations.
Detection of novel situations, where the robot knowledge is not sufficient, plays then an important
role for the subsequent adapting and learning of new concepts and its this thesis main topic.

Additionally in the context of mobile robotics, spatial concepts and semantics play a main role to perform
complex human-like tasks and human interactions. For handling those, a robot builds a representation
of space extended with semantic properties, process which is known as semantic mapping.
This representation identifies spatial entities and classifies them according to their meaning to
humans allowing the robot to reason at a very high-level.
For example, in a room scale, humans categorize space in kitchens, bedrooms, corridors,
office, theater, computer lab, etc\dots, by relating room areas with those categories, that describing expected
properties and actions that take place on them, a robot can solve tasks like:
``Where are the cornflakes?''

\section{Problem and Goals}
The semantic mapping process plays an important role for allowing agents to reason at an high-level
and is totally defined by the knowledge the agent has on the spatial concepts and semantic categories.
For that reason there is an high interest in creating methods that are able to alert the agent for
gaps on its knowledge.
And with that purpose, this thesis, studies novelty detection techniques and how to apply them to
detect novel room categories during a semantic mapping process.
To this end, the semantic mapping process proposed by \cite{pronobis2011phd} is studied and a method
to detect novel categories using graphical models is suggested.

\section{Semantic Mapping}
Semantic mapping is the process by which an agent builds a representation of the environment and
enhances the identified entities with semantic values (i.e.\ human semantics).
In this thesis the semantic mapping process proposed by \cite{pronobis2011phd} is used as base.

\subsection{Spatial Knowledge Representation}
In order to perform semantic mapping an agent needs to possess knowledge on how to segment and
identify entities existent in space by analysing the sensed data. For that
Pronobis~\cite{pronobis2011semmap} presents a layered knowledge structure:
the \emph{sensory layer} maintains an accurate and exact description of environment in a machine-friendly representation,
the \emph{place layer} employs knowledge on how to segment and identify areas such as rooms,
the \emph{categorical layer} defines knowledge on properties and categories that allow to extract
higher-level information from the sensed data
and the \emph{conceptual layer} maintains knowledge on entities concepts, the relations between them
and information on how to map them to human semantics.

\subsection{Conceptual Map}
The semantic mapping is then based on the instantiation of a conceptual map that uses all the
ontology defined on the conceptual layer in order to build a probabilistic graphical model that
relates and propagates probabilities across all the entities and properties detected by the
other layers.

By using probabilistic graphical models, the conceptual map becomes then a generative model that
allows to perform inference on variable configurations. For that it is used as a base tool
for novelty detection of unknown categories on some of its variables.

\section{Novelty Detection}
Novelty detection, also known as outlier or anomaly detection, is a
classification problem related to identification of new or unknown data
patterns that the system is not aware of~\cite{markou2003novelty}.
The ability to identify novel cases is crucial in any autonomous system
that is deployed to unknown or to uncontrolled environments, as it gives the
system the ability to detect that something is not conforming to its knowledge and
therefore should be treated with caution.
% It has several applications such as fault detection~\cite{tarassenko1999novelty},
% intrusion detection~\cite{fan2001using},
% detection of masses in mammograms~\cite{tarassenko1995novelty} or detection of
% novel and useful documents~\cite{zhang2002novelty}.

\subsection{Novelty Detection as Threshold}
Due to the noisy, uncertain and insufficient nature of sensed data, novelty ought to be
treated in a probabilistic way in where each sample has a certain probability
of being generated by a class not known for the agent and a complementary probability $P(\overline{novel}|x)$
of being generated by a known class.

Additionally any detector can be uniquely described by the set $N$ of samples that it
classifies as novel. With that its possible to show that by including a new sample $a$ in $N$ an agent
can increase its detection rate at a cost of producing more mis-classifications.
This describes the base of the \emph{error and rejection tradeoff} described by \cite{chow1970optimum}.
Due to the interest of having the highest detection rate with a given minimum error rate, any optimal
detector can be described as a \emph{continuous knapsack problem} of the sample space.
The greedy ordering defined to solve it can be shown to be equivalent to:
\begin{eqnarray}
P(\overline{novel}|x) &=& \frac{P(x|\overline{novel})P(\overline{novel})}{P(x)}
\end{eqnarray}

\subsection{Assumption on a constant $P(novel)$}
By considering $P(novel)$ to be constant a ratio between conditional and unconditional probabilities
of the sensed variables is obtained.
Although it sounds reasonable to assume the probability of specific room belong to a novel
category is constant, this thesis points that this assumption might be very strong and not
reasonable in a realistic scenario. The main point being that in this case the graph structure
changes between samples and some might be more prone to produce novel samples, as opposed to normal
classification scenarios where its clear that the category is indeed the only generator of features.

\subsection{Modeling Conditional Probability}
Since the graphical model produced by the conceptual map tries to model the distribution of
variables assuming the knowledge of the agent holds true, it represents a method to model
$P(x|\overline{novel})$.
%since it allows to calculate the density probability that the set of features $x$ is
%sensed given that all the variables and graph structure, including the category of
%a specific room $a$ are correctly modelled by the agent knowledge.
%This way the distribution modelled by a factor graph equivalent to the chain-graph used by
%the conceptual map can be used as an approximation for the conditional probability of $x$.

\subsection{Modeling Unconditional Probability}
With no knowledge on the real distribution samples, the correct approach is to model it
with a uniform distribution. Nonetheless in the presented case, it was assumed that only the room
category of a specific room is unknown and that the remaining knowledge holds trues.
In that case, the correct approach is to replace the variable that represents the room where to
detect novelty with a single factor connecting all variables directly dependent on it.

Due to its high-dimensionality, approximating that factor with unlabelled data is infeasible and
approximations have to be done. In this thesis two approximations were presented: an uniform model
and an independent model. In the independent model, the
single\hyp{}connected factors are trained from unlabelled data and its expected that they create a better
approximation to the unconditional probability than an assumption on the uniform model.

\section{Results}
A synthetic dataset modelling the expected features when implementing on the real case and enforcing
a constant $P(novel)$ was used for testing the developed models.
The two suggested models where compared between them and to an optimal case.
The tests across several situations shows how they behaviour change as more information is sensed
from the environment.

@@

\section{Conclusions}
This thesis studied the problem of novelty detection of semantic categories on the spatial knowledge
of mobile robots. To that purpose a semantic mapping process~\cite{pronobis2011semmap} was studied and
a novelty detection scheme based on the conceptual map used by it proposed.
Initial tests where performed on synthetic data showing an interest in using unlabelled data to
increase the detection performance.
Additionally all the assumptions and bases behind novelty detection were presented and several
directions on how to create a generalized framework for novelty detection were discussed.

>>>>

\section{Motivation}
For a long time humanity has dreamed that one day robots will be among us. They will explore our world
and interact with us, understand our concepts and reason. An important step in that direction is endowing 
robots with knowledge about human concepts and semantics. However, it is unrealistic to believe that the 
human world can be fully modeled in the robot's brain at the design stage. Therefore, robots must be able 
to adapt and learn when confronted with novel situations. Detection of novel situations, where the knowledge
of the robot is not sufficient plays an important role in adaptation and learning of new concepts and 
is the main topic of this thesis.

In the context of mobile robotics, spatial concepts and semantics are crucial to enable the robot to perform
complex human-like tasks and human interactions. For handling those, a robot builds a representation
of space extended with semantic properties, process which is known as semantic mapping.
This representation identifies spatial entities and classifies them according to their meaning to
humans allowing the robot to reason at a high abstraction level. For example, humans categorize spaces as 
kitchens, bedrooms, corridors, offices, theaters, computer labs, etc. based on spatial properties, 
objects and actions that are characteristic of those spaces. Similarly, if a robot had a knowledge about
those spatial concepts, it could answer such questions as: ``Where are cornflakes?''

\section{Problem and Goals}
The semantic mapping process plays an important role for allowing artificial mobile agents to reason at 
a high level and is dependent on the knowledge that the agent has about the spatial concepts and semantic 
categories. For that reason there is high interest in creating methods that are capable of detecting 
gaps in the agent's knowledge. For that purpose, this thesis, studies novelty detection techniques and 
how they can be applied to detection novel room categories during the semantic mapping process.
To this end, the semantic mapping process proposed in \cite{pronobis2011phd} is studied and a method
to detect novel categories using graphical models is suggested.

\section{Semantic Mapping}
Semantic mapping is the process by which an agent builds a representation of the environment and
augments the identified spatial entities with semantic values (i.e.\ human semantics).
In this thesis the semantic mapping process proposed in \cite{pronobis2011phd} is used as base.

\subsection{Spatial Knowledge Representation}
In order to perform semantic mapping an agent needs to possess knowledge on how to segment and
identify entities existent in space by analysing the sensed data. For that, 
Pronobis~\cite{pronobis2011semmap} presents a spatial knowledge representation consisting of four layers:
the \emph{sensory layer} maintains an accurate and exact description of the environment in a machine-friendly 
representation, the \emph{place layer} represents the world in terms of a set of discrete places and connectivity
between them, the \emph{categorical layer} defines knowledge about properties and categories that allow to extract
higher-level information from the sensed data, and the \emph{conceptual layer} maintains knowledge about 
spatial entities and concepts as well as the relations between them.

\subsection{Conceptual Map}
The semantic mapping is based on the instantiation of a conceptual map that uses the
ontology defined in the conceptual layer in order to build a probabilistic graphical model that
relates and propagates probabilities across all the entities and properties detected by the
other layers.

By using probabilistic graphical models, the conceptual map becomes a generative model that
permits inference about different variables. This enables the use of the conceptual map 
as a tool for detection of unknown categories.

\section{Novelty Detection}
Novelty detection, also known as outlier or anomaly detection, is a
classification problem related to identification of new or unknown data
patterns that the system is not aware of~\cite{markou2003novelty}.
The ability to identify novel cases is crucial in any autonomous system
that is deployed in unknown or uncontrolled environments, as it permits 
detection of cases that do not conform to the robot's knowledge and
therefore should be treated with caution.
% It has several applications such as fault detection~\cite{tarassenko1999novelty},
% intrusion detection~\cite{fan2001using},
% detection of masses in mammograms~\cite{tarassenko1995novelty} or detection of
% novel and useful documents~\cite{zhang2002novelty}.

\subsection{Novelty Detection as Thresholding}
Due to the nature of the sensed data which are noisy and uncertain, novelty ought to be
treated in a probabilistic way where each sample has certain probability
of being generated by a class not known to the agent and a complementary probability $P(\overline{novel}|x)$
of being generated by a known class.

Additionally, any detector can be uniquely described by the set $N$ of samples that it
classifies as novel. With that, it is possible to show that by including a new sample $a$ into 
$N$, an agent can increase its detection rate at a cost of producing more misclassifications.
This describes the base of the \emph{error and rejection tradeoff} described in \cite{chow1970optimum}.
Due to the goal of having the highest detection rate with a given minimum error rate, any optimal
detector can be described as a \emph{continuous knapsack problem} of the sample space.
The resulting greedy ordering can be shown to be equivalent to:
\begin{eqnarray}
P(\overline{novel}|x) &=& \frac{P(x|\overline{novel})P(\overline{novel})}{P(x)}
\end{eqnarray}

\subsection{Assumption about Constant $P(novel)$}
By considering $P(novel)$ to be constant, a ratio between conditional and unconditional probabilities
of the sensed variables is obtained. Although it sounds reasonable to assume that the probability of a specific 
room belonging to a novel category is constant, this thesis points out that this assumption might be very strong 
and not reasonable in realistic scenarios. The main problem is related to the fact that the graph structure
changes as the robot gathers new samples and some structures might be more prone to produce novel samples. This
is different from the normal classification scenarios where its clear that the category is indeed the only 
generator of features.

\subsection{Modeling Conditional Probability}
Since the graphical model produced by the conceptual map tries to model the distribution of
variables assuming that the knowledge of the agent is representative of the world, it models $P(x|\overline{novel})$.
%since it allows to calculate the density probability that the set of features $x$ is
%sensed given that all the variables and graph structure, including the category of
%a specific room $a$ are correctly modelled by the agent knowledge.
%This way the distribution modelled by a factor graph equivalent to the chain-graph used by
%the conceptual map can be used as an approximation for the conditional probability of $x$.

\subsection{Modeling Unconditional Probability}
With no knowledge about the real distribution of samples, the correct approach is to model it
with a uniform distribution. Nonetheless in the presented case, it was assumed that only the room
category of a specific room is unknown and that the remaining knowledge holds true.
In that case, the correct approach is to replace the variable that represents the room where the 
novelty is detected with a single factor connecting all variables directly dependent on it.

Due to its high-dimensionality, approximating that factor with unlabelled data is infeasible and
approximations have to be done. In this thesis two approximations were presented: a uniform model
and an independent model. In the independent model, the single\hyp{}connected factors are trained from 
unlabelled data and its expected that they create a better approximation to the unconditional probability 
than an assumption about the uniform model.

\section{Results}
A synthetic dataset modelling the features expected when deployed in a real scenario and enforcing
a constant $P(novel)$ was used for testing the developed models. The two suggested models where compared 
to each other and to an optimal case. The tests across several situations show how the behaviour changes 
as more information is sensed from the environment.

@@

\section{Conclusions}
This thesis studied the problem of detection of novel semantic categories of rooms in the context of
semantic mapping with mobile robots. To that purpose, a semantic mapping process~\cite{pronobis2011semmap} 
was studied and a novelty detection scheme based on a conceptual map was proposed.
Initial tests where performed on synthetic data showing benefit of using unlabelled data to
increase the detection performance. Additionally, all the assumptions behind novelty detection were 
presented and several directions on how to create a generalized framework for novelty detection 
were discussed.
